---
title: "Xaringan Themes"
description: |
  Themes for the Xaringan Slide Show Format
author:
  - name: Andy Grogan-Kaylor 
    url: https://agrogan1.github.io/
    affiliation: University of Michigan
    affiliation_url: https://www.umich.edu
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    highlight: haddock
    toc: yes
    code_folding: true
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE)

options(htmltools.dir.version = FALSE, 
        htmltools.preserve.raw = FALSE)

library(xaringanExtra)

```

# Introduction

[Xaringan](https://github.com/yihui/xaringan) is a slideshow system for RMarkdown. Alison Hill has an [excellent demonstration of the possibilities of Xaringan](https://arm.rbind.io/slides/xaringan.html).

# Themes

## University of Michigan Xaringan Theme

```{r}

embed_xaringan(url = "https://agrogan1.github.io/R/xaringan-themes/MichiganTest.html",
               ratio = "4:3")

```

* [RMarkdown Document](https://github.com/agrogan1/R/blob/master/xaringan-themes/MichiganTest.Rmd)
* [michigan.css](https://github.com/agrogan1/R/blob/master/xaringan-themes/michigan.css)
* [michigan-fonts.css](https://github.com/agrogan1/R/blob/master/xaringan-themes/michigan-fonts.css)

## OsloMet Xaringan Theme

```{r}

embed_xaringan(url = "https://agrogan1.github.io/R/xaringan-themes/OsloMetTest.html",
               ratio = "4:3")

```

Inspired by: [https://github.com/martinhelso/OsloMet](https://github.com/martinhelso/OsloMet)

* [RMarkdown Document](https://github.com/agrogan1/R/blob/master/xaringan-themes/OsloMetTest.Rmd)
* [oslomet.css](https://github.com/agrogan1/R/blob/master/xaringan-themes/oslomet.css)
* [oslomet-fonts.css](https://github.com/agrogan1/R/blob/master/xaringan-themes/oslomet-fonts.css)

## Tufte(-ish) Xaringan Theme

```{r}

embed_xaringan(url = "https://agrogan1.github.io/R/xaringan-themes/TufteTest.html",
               ratio = "4:3")

```

* [RMarkdown Document](https://github.com/agrogan1/R/blob/master/xaringan-themes/TufteTest.Rmd)
* [tufte.css](https://github.com/agrogan1/R/blob/master/xaringan-themes/tufte.css)
* [tufte-fonts.css](https://github.com/agrogan1/R/blob/master/xaringan-themes/tufte-fonts.css)

## *Moments of Rising Mist* Xaringan Theme

```{r}

embed_xaringan(url = "https://agrogan1.github.io/R/xaringan-themes/Moments-Of-Rising-Mist.html",
               ratio = "4:3")

```

CSS to replicate design of *Moments of Rising Mist: A Book of Sung Poetry*

* [RMarkdown Document](https://github.com/agrogan1/R/blob/master/xaringan-themes/Moments-Of-Rising-Mist.Rmd)
* [moments-of-rising-mist.css](https://github.com/agrogan1/R/blob/master/xaringan-themes/moments-of-rising-mist.css)
* [moments-of-rising-mist-fonts.css](https://github.com/agrogan1/R/blob/master/xaringan-themes/moments-of-rising-mist-fonts.css)

## Banff Xaringan Theme

```{r}

embed_xaringan(url = "https://agrogan1.github.io/R/xaringan-themes/BanffTest.html",
               ratio = "4:3")

```

Based upon design ideas from the [Banff Mountain Film Festival](https://www.banffcentre.ca/banffmountainfestival/tour)

* [RMarkdown Document](https://github.com/agrogan1/R/blob/master/xaringan-themes/BanffTest.Rmd)
* [banff.css](https://github.com/agrogan1/R/blob/master/xaringan-themes/banff.css)
* [banff-fonts.css](https://github.com/agrogan1/R/blob/master/xaringan-themes/banff-fonts.css)

# Supplemental Materials

* My page on HTML Slides: [https://agrogan1.github.io/myposts/slide-formats.html](https://agrogan1.github.io/myposts/slide-formats.html)
* [Branding Guidelines from University of Michigan](https://brand.umich.edu/)
* Xaringan: [https://github.com/yihui/xaringan](https://github.com/yihui/xaringan)
* Xaringan Themes: [https://github.com/yihui/xaringan/wiki/Themes](https://github.com/yihui/xaringan/wiki/Themes)

# Example

* My [ggplot flipbook](../ggplot-flipbook/ggplot-flipbook.html) uses the  CSS from the University of Michigan Xaringan theme.

# Possibly Related

> Possibly related, and possibly helpful, may be my R Package for [Michigan Colors in R Graphs](https://agrogan1.github.io/michigancolors/)







