---
title: "OsloMet Test"
author: "Andy Grogan-Kaylor"
institute: "University of Michigan"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [oslomet.css, oslomet-fonts.css]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}

options(htmltools.dir.version = FALSE)

```

```{r, echo=FALSE}

xaringanExtra::use_tile_view()

xaringanExtra::use_share_again()

xaringanExtra::use_progress_bar(color = "#FFDC00", 
                                location = "bottom")


```

# Introduction

Following the inspiration of the OsloMet Beamer theme at: [https://github.com/martinhelso/OsloMet](https://github.com/martinhelso/OsloMet)

.important-message-box[Important 

Press `o` for an overview of the presentation

]

---
# Bulleted List

* Item 1

--

* Item 2

--

* Item 3

---
# Highlighting

This text is .highlight[highlighted].

This text is a .description[description].

---
# Theorem

.theorem-box[The first line is considered to be the theorem header.

This is the theorem content. 

$a^2 + b^2 = c^2$

]

---
# Example

.example-box[Example

This is the content of the example.

]

---
# Important Message

.important-message-box[Important 

The content of this important message goes here.

]

---
# Blockquote

> This is a blockquote.






