---
title: "A Two Page Guide to R"
author: "Andy Grogan-Kaylor"
date: '`r format(Sys.Date(), format="%B %d, %Y")`'
output:
  tint::tintPdf:
    number_sections: yes
    toc: no
    highlight: haddock
  tufte::tufte_handout:
    number_sections: yes
    toc: no
  tint::tintHtml:
    number_sections: yes
    toc: yes
    highlight: haddock
  pdf_document:
    fig_height: 2.5
    highlight: haddock
    latex_engine: xelatex
  html_document:
    fig_height: 3
    highlight: haddock
    keep_md: no
    number_sections: yes
    theme: default
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: yes
mainfont: Arial
geometry: margin=0.25in
subtitle: For Applied Researchers and Program Evaluators
fontsize: 10pt
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      fig.margin = TRUE)

```

```{r, echo = FALSE}

library(DiagrammeR)

library(pander)

library(tufte)

```

# Why?[^difficult]

[^difficult]: R has a reputation for being difficult to learn, and a lot of that reputation is deserved.  

[R](https://www.r-project.org/) is open source, and therefore free, statistical software that is particularly good at obtaining, analyzing and visualizing data.[^scriptsanddata]

[^scriptsanddata]: R Commands are stored in a *script* or *code* file that usually ends in .R, e.g. `myRscript.R`. The command file is distinct from your actual data, stored in an .RData file, e.g. `mydata.RData`.

I believe that it is possible to teach R in an accessible way, and that **a little bit of R can take you a long way**.

A great deal of data analysis and visualization involves the same core set of steps.[^coresteps]

[^coresteps]: Given the fact that we often want to apply the same core set of tasks to new questions and new data, there are ways to overcome the steep learning curve and learn a replicable set of commands that can be applied to problem after problem. **The same 5 to 10 lines of R code can often be tweaked over and over again for multiple projects.**

$$\text{have a question} \rightarrow \text{get data} \rightarrow \text{process and clean data} \rightarrow$$ 
$$\text{visualize data} \rightarrow \text{analyze data} \rightarrow \text{make conclusions}$$

# Get R[^getR]

[^getR]: [R](https://www.r-project.org/) is available at [https://www.r-project.org/](https://www.r-project.org/). R is a lot easier to run if you run it from RStudio, [http://www.rstudio.com](http://www.rstudio.com).

# Get Data[^getdata]

[^getdata]: Data often comes from other types of data files like SPSS, Stata, or Excel.  Especially in beginning R programming, getting the data into R can be the most complicated part of your program.

```{r, eval=FALSE, echo=TRUE}

load("the/path/to/mydata.Rdata") # data in R format

library(foreign) # library for importing data 
mydata <- read.spss("the/path/to/mySPSSfile.sav") # SPSS
mydata <- read.dta("the/path/to/myStatafile.dta") # Stata

library(readxl) # library for importing Excel files
mydata <- read_excel("the/path/to/mySpreadsheet.xls")

save(mydata, file = "mydata.RData") # save in R format

```

```{r, eval = TRUE, echo = FALSE}

N <- 100

x <- rnorm(N, 100, 10)

e <- rnorm(N, 0, 1 )

y <- x + e

z <- rbinom(N, 1, .25) # binary variable

mydata <- data.frame(x,y,z)

```

```{marginfigure}

R makes a strong distinction between *continuous* *numeric* variables that measure  scales like mental health or neighborhood safety, and *categorical* *factor variables* that measure non-ordered categories like religious identity or gender identity. Many statistical and graphical procedures are designed to recognize and work with different variable types.  

```

# Manage Data [^dollarsign] 

```{r}

mydata$x[mydata$x == -9] <- NA # missing to NA

```

```{r, tidy=FALSE}

mydata$w <- factor(mydata$z, # original numeric variable
                   levels = c(0, 1, 2), 
                   labels = c("Group A", "Group B", "Group C"), 
                   ordered = TRUE) # whether order matters

mydata$z <- as.numeric(mydata$w) # factor / character to numeric

```

```{marginfigure}

You *don't* need to use all of the options.  e.g. `mydata$w <- factor(mydata$z)` will work just fine.

**Changing variables from factor to numeric, and vice versa can sometimes be the simple solution that solves a lot of problems when you are trying to graph your variables.**

```

# Visualize Data

## Histogram

```{r, tidy=FALSE}

hist(mydata$x, # what I'm graphing
     main = "your title goes here", # title
     xlab = "income", # label for x axis
     col = "blue") # color

```

```{marginfigure}

You *don't* need to use all of the options.  e.g. `hist(mydata$x)` will work just fine.

```

## Barplot

```{r, tidy=FALSE}

barplot(table(mydata$z), # what I'm graphing
        main = "your title goes here", # title
        xlab = "group", # label for x axis
        col = "gold") # color

```

```{marginfigure}

You *don't* need to use all of the options.  e.g. `barplot(table(mydata$z))` will work just fine.

```

## Scatterplot

```{r, tidy=FALSE}

plot(mydata$x,mydata$y, # plot x and y
     main = "your title goes here", # title
     xlab = "income", # label for x axis
     ylab = "mental health", # label for y axis
     pch = 19, # Plot CHaracter, 19 is filled dots
     las = 2, # LAbel Style, 2 is "perpendicular"
     col = "darkgreen") # color

```


```{marginfigure}

You *don't* need to use all of the options.  e.g. `plot(mydata$x,mydata$y)` will work just fine.

```

```{marginfigure}

When scatterplots have fewer dots than you think they should have, often due to "over-printing", adding some random noise, or "jittering" the dots in the scatterplot may help:  `plot(jitter(mydata$y, factor = 5000) ~ mydata$x)`.  Experiment with different sizes of *factor*.

```

# Descriptive Statistics

```{marginfigure}

For another approach to summarizing your data, try `library(skimr)` then `skim(mydata)` or `skim(mydata$x)`.

```

```{r}

summary(mydata$x) # for continuous or factor variables

table(mydata$z) # especially suitable for factor variables

```

# Credits

Prepared by [Andy Grogan-Kaylor](https://agroganweb.wordpress.com/)

```{r, echo=FALSE,eval=FALSE}

## t-Tests and Correlations[^ttest]

# [^ttest]: A t-test is a test of whether there are differences between two groups.  A correlation is a measure, varying between -1 and +1 of the relationship between two continuous variables.

```

```{r, echo = FALSE, eval = FALSE}

t.test(mydata$y ~ mydata$z) # y is numeric; z is a factor

cor(mydata$x, mydata$y) # get the correlation

```

```{r, echo=FALSE,eval=FALSE}

## Regressions[^regression]

[^regression]: A regression is an analysis of the influence of one or more independent variables on an outcome of interest.

```

```{r, echo = FALSE, eval = FALSE}

lm(y ~ x, data = mydata) # regression of y on x

```

[^dollarsign]:  The \$ sign is a kind of "connector".  mydata\$x means:  "The variable x in the dataset called mydata".


