---
title: "Multi-Page dplyr"
author: "Andy Grogan-Kaylor"
date: "`r Sys.Date()`"
output: 
  tint::tintPdf:
    number_sections: yes
    toc: yes
    highlight: haddock
  tint::tintHtml:
    number_sections: yes
    toc: yes
    highlight: haddock
geometry: margin=0.5in
mainfont: Roboto Condensed
fontsize: 12pt
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      fig.margin = TRUE)

library(tibble)

library(pander)

```

# Background

`dplyr` is a very powerful R library for managing and processing data.[^dplyr]

[^dplyr]: The origins of the name `dplyr` seem somewhat obscure, but I sometimes think of this package as the *data plyers*.

While `dplyr` is very powerful, learning to use `dplyr` can be very confusing. This guide aims to present some of the most common `dplyr` functions and commands in the form of a brief cheatsheet.

```{r}

library(dplyr)

```

# Simulated Data

```{r, echo=FALSE}

N <- 5

year <- round(runif(N, min = 2015, max = 2018),
              digits = 0)


x <- round(rnorm(N, 40, 10), digits = 2)

x[1] <- NA # introduce a missing value

y <- rbinom(N, 1, .25)

y[2] <- 2 # introduce alternate value

y <- factor(y,
            levels = c(0, 1, 2),
            labels = c("Group A", "Group B", "Group C"))

z <- round(rnorm(N, 100, 10), digits = 2)

mydata <- tibble(year, x, y, z)

```

```{r, echo=FALSE}

pander(head(mydata))
  
```

# Piping

*Pipes* `%>%` connect pieces of a command e.g. *data* to *data wrangling* to a *graph command.*

`dplyr` commands will often look something like the outline below.

```{r, eval=FALSE}

mydata %>% 
  data_wrangling %>% 
  more_data_wrangling %>%
  graph_command

```


# Aggregate Data: `group_by()` & `summarise()`

```{r}

mynewdata <- mydata %>% 
  group_by(year) %>% # group by y
  summarise(mean_x = mean(x), # mean of x
            n = n()) # count up

```

```{r, echo=FALSE}

pander(head(mynewdata))
  
```

# Select A Subset of Variables: `select()`

```{r}

mynewdata <- mydata %>% 
  select(x,y) # select only x and y

```

```{r, echo=FALSE}

pander(head(mynewdata))
  
```

# Filter A Subset of Rows: `filter()`

```{r}

mynewdata <- mydata %>% 
  filter(year > 2010) # filter on year

```

```{r, echo=FALSE}

pander(head(mynewdata))
  
```

# Create New Variables: `mutate()`

```{r}

mynewdata <- mydata %>% 
  mutate(myscale = x + z) # create a new variable e.g. a scale

```

```{r, echo=FALSE}

pander(head(mynewdata))
  
```

# Recode Variables: `mutate()`

## Continuous Into Categorical: `mutate()` & `cut()`

```{r}

mynewdata <- mydata %>% 
  mutate(zcategorical = cut(z, # cut at breaks
                            breaks=c(-Inf, 100, Inf), 
               labels = c("low", "high")))

```

```{r, echo=FALSE}

pander(head(mynewdata))
  
```

## Categorical Into Categorical: `mutate()` & `recode()`

```{r}

mynewdata <- mydata %>% 
  mutate(yrecoded = dplyr::recode(y, # recode values
                         "Group A" = "Red Group",
                         "Group B" = "Blue Group", 
                         .default = "Other"))
```

```{r, echo=FALSE}

pander(head(mynewdata))
  
```

# Rename Variables: `rename()`

```{r}

newdata <- mydata %>% 
  rename(age = x, # rename
         mental_health = z)

```

```{r, echo = FALSE}

pander(head(newdata))

```

# Drop Missing Values: `filter()`

```{r}

newdata <- mydata %>% 
  filter(!is.na(x)) # filter by x is not missing

```

```{r, echo = FALSE}

pander(head(newdata))

```

# Random Sample

```{r}

newdata <- mydata %>% 
  sample_frac(.5) # fraction of data to sample

```

```{r, echo = FALSE}

pander(head(newdata))

```

# Connecting To Other Packages Like `ggplot`

Notice how, in the code below, I never actually create the new data set `mynewdata`.  I simply pipe `mydata` into a `dplyr` command, and pipe the result directly to `ggplot2`.

```{r}

library(ggplot2)

mydata %>% # my data
  mutate(myscale = x + z) %>% # dplyr command to make new variable 
  filter(y != "Group C") %>% # filter on values of y
  ggplot(aes(x = year, # the rest is ggplot
             y = myscale)) +
  geom_point() + # points
  geom_smooth(se = FALSE, # smoother without confidence interval
              method = "lm") + # linear smoother
  labs(title = "My Scale By Year") + # labels
  theme(axis.text.x = element_text(size = 10, # tweak theme
                                   angle = 90))
  
```
















